\bf A note on the instability and degeneracy of deep learning models

### 2017-11-02 writing/paper-note/docs/note_blinded.tex
dr. nordman's changes

\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{longtable,booktabs}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={A Note on the Instability and Degeneracy of Deep Learning Models},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{5}

@@

\title{\bf A Note on the Instability and Degeneracy of Deep Learning Models}
\author{Andee Kaplan \\ Department of Statistics, Iowa State University \\  and \\ Daniel Nordman \\ Department of Statistics, Iowa State University \\  and \\ Stephen Vardeman \\ Department of Statistics and Department of Industrial and\\
Manufacturing Systems Engineering, Iowa State University \\ }

@@

We consider the behavior, and the potential impropriety, of sequences of
discrete probability models built to incorporate observations of
increasing sample size \(N\). Interest is in identifying instability in
such models, which is roughly characterized by probabilities with
extreme sensitivity to small changes in data configuration. The concept
of instability was introduced in the field of statistical physics (i.e.,
point processes) by Ruelle
(\protect\hyperlink{ref-ruelle1999statistical}{1999}) and then further
extended by Schweinberger Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) for a
family of exponential models. At issue, models exhibiting instability
are typically undesirable as these tend to provide poor representations
of data or data-generation. As an example, such models can include
near-degenerate distributions that assign all essential probability mass
to only a subset of an overall sample space. The latter issue in
connection to degeneracy has been recognized as a concern in that
dominant model outcomes may not resemble observed data (cf. Handcock
\protect\hyperlink{ref-handcock2003assessing}{2003}). As a compounding
issue, model instability often has direct negative impacts for
statistical inference and computations based on likelihood functions.
Namely, volatilities in probability structure can potentially hamper the
numerical evaluations required for maximum likelihood estimation as well
as other model-based simulations via Markov Chain Monte Carlo (MCMC).
These reasons motivate our general study of instability for a broad
class of probability models, described next.

@@

For random variables
\(\boldsymbol X \equiv\boldsymbol X_N= (X_1, \dots, X_N)\) with sample
space \(\mathcal{X}^N\), \(|\mathcal{X}| < \infty\), consider an
exponential family model for \(\boldsymbol X\) with probability mass
function given by
\begin{equation}
\label{eq:expo}
p_{N, \boldsymbol \theta}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) - \psi(\boldsymbol \theta)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
\end{equation}
depending on parameter vector
\(\boldsymbol \theta \in \Theta_N \subset \mathbb{R}^{k}\) and natural
parameter function
\(\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L\) with fixed
positive integers \(k\) and \(L\) denoting their dimensions. Above,
\(\boldsymbol g_N : \mathcal{X}^N \mapsto \mathbb{R}^L\) is a vector of
sufficient statistics, while \[
\psi(\boldsymbol \theta) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) \right], \qquad \boldsymbol \theta \in \Theta_N\equiv \{\boldsymbol \theta \in \mathbb{R}^k : \psi(\boldsymbol \theta) < \infty \},
\] denotes the normalizing function with parameter space \(\Theta_N\).
The natural parameter function \(\eta (\boldsymbol \theta)\) has a
linear form (i.e.,
\(\eta (\boldsymbol \theta)= \bm{A} \boldsymbol \theta\) for a given
\(L \times k\) matrix \(\bm{A}\)) in many common model formulaitons,
though may also be nonlinear (e.g., curved exponential families). In the
linear case, \(\eta (\boldsymbol \theta) = \boldsymbol \theta\) may be
generally assumed in the exponential parameterization with a minor
modification to the definition of sufficient statistics
\(\boldsymbol g_N(\boldsymbol x)\).

@@

For model sequences
\(P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \theta_N}(\boldsymbol x)\)
of the exponential type \eqref{eq:expo}, such as those in
\eqref{eq:mod1}-\eqref{eq:mod2}, note that the dimension \(k\) of the
parameter \(\boldsymbol \theta_N\in\Theta \subset \mathbb{R}^k\)
necessarily remains the same for all sample sizes \(N \geq 1\) as the
form of the natural parameter function \(\eta(\cdot)\) in \eqref{eq:expo}
and the number of sufficient statistics
\(\boldsymbol g_{N}(\boldsymbol x)\) do not depend on \(N\).
Consequently, \(\boldsymbol \theta_N\) lies in a parameter space of
fixed Euclidean dimension \(k\). However, this aspect need not be true
for other types of FOES models considered in Sections
\ref{restricted-boltzmann-machines} - \ref{deep-learning}, where instead
the numbers of parameters and sufficient statistics commonly increase
with the sample size \(N\).

\subsection{Restricted Boltzmann
Machines}\label{restricted-boltzmann-machines}

@@

This definition of S-unstable is a generalization or reinterpretation of
``unstable'' used in Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) by allowing
possibly non-exponential family models (e.g., RBM and DBM models in
Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning} as well
as a potentially increasing number \(q(N)\) of parameters through the
parameter sequence \(\boldsymbol \theta_N\in \mathbb{R}^{q(N)}\). While
this definition differs in form and scope from the original, it does
match that in Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) for the
special case of exponential models (cf.~Section
\ref{discrete-exponential-family-models} considered there. Section
\ref{illustrations} provides several examples of unstable models as well
as causes for model instability, where the latter may often be traced to
issues in model form (i.e., data functions) and/or parameterization. We
next describe several potentially undesirable features associated with
S-unstable FOES models.

\BeginKnitrBlock{remark}
\iffalse <span class="remark"><em>Remark. \fi{} In the definition
\eqref{eq:Sun} of S-instability, we note that the numerical measure
\(\REP(\boldsymbol \theta_N)/N\) of model instability is invariant to
\emph{independent replications} of data. That is, let \(M \geq 1\)
denote a possible number of replications and consider data
\(\boldsymbol Y_{N,M} \equiv (\boldsymbol X^{(1)}_N, \dots, \boldsymbol X^{(M)}_N)\)
formed by \(\{ \boldsymbol X^{(j)}_N\}_{j=1}^M\) as \(M\) iid
replications of a random vector \(\boldsymbol X_N=(X_1,\ldots,X_N)\),
where the latter follows a FOES model with probabilities
\(P_{ \boldsymbol \theta_N}(\boldsymbol x)>0\),
\(\boldsymbol x\in\mathcal{X}^N\). This leads to a joint model, say
\(P_{ \boldsymbol \theta_N}(\boldsymbol y)\),
\(\boldsymbol y\in\mathcal{X}^{NM}\), for \(\boldsymbol Y_{N,M}\)
consisting of \(N*M\) random variables in total. Then, the LREP for
\(\boldsymbol Y_{N,M}\), scaled by associated size, is given by
\begin{align*}
\frac{1}{NM}\REP_{\boldsymbol Y_{N,M}}( \boldsymbol\theta_N ) &\equiv \frac{1}{NM}\log\left[\frac{\max_{\boldsymbol y \in \mathcal{X}^{NM}}P_{ \boldsymbol \theta_N}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^{NM}}P_{ \boldsymbol \theta_N}(\boldsymbol y)} \right] \\
&= \frac{1}{NM}\log\left[\frac{\max_{\boldsymbol x \in \mathcal{X}^{N}}P_{ \boldsymbol \theta_N}(\boldsymbol x)  }{\min_{\boldsymbol x \in \mathcal{X}^{N}}P_{ \boldsymbol \theta_N}(\boldsymbol x)} \right]^M \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ),
\end{align*}
where
\(\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ) \equiv \REP( \boldsymbol\theta_N )\)
denotes the log-ratio of extremal probabilities for \(\boldsymbol X_N\)
defined from \eqref{eq:elpr}. That is, due to iid properties, the
sample-size corrected LREP for \(\boldsymbol Y_{N,M}\) equals the
analog, \(\REP(\boldsymbol\theta_N )/N\), from the underlying common
data model for \(\boldsymbol X_N\) alone, regardless of the level
\(M \geq 1\) of independent replication. Consequently, the definition of
an S-unstable model is unaffected by independent replication and all
instability properties may be characterized by those of one observation
from the common FOES model. For computational purposes, this aspect also
implies that if the original data \(\boldsymbol X_N=(X_1,\ldots,X_N)\)
in a FOES model consist of \(N\) iid random variables, then the
size-scaled log-ratio \eqref{eq:elpr} may be calculated as \[
\frac{1}{N}\REP( \boldsymbol\theta_N ) \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ) = \log\left[ \frac{\max_{ x \in \mathcal{X}}P_{\boldsymbol\theta_N}(X_1=x)} {\min_{ x \in \mathcal{X}}P_{\boldsymbol\theta_N}(X_1=x)} \right]
\] based on the extremal probabilities of just one random variable
\(X_1\).
\EndKnitrBlock{remark}

@@

The above model condition incorporates many standard parameterizations
and follows, for instance, whenever
\(P_{\boldsymbol \theta_N}(\boldsymbol x)/P_{\boldsymbol \theta_N}(\boldsymbol y) = [P_{-\boldsymbol \theta_N}(\boldsymbol x)/P_{-\boldsymbol \theta_N}(\boldsymbol y)]^{-1}\)
holds for outcomes \(\boldsymbol x, \boldsymbol y \in\mathcal{X}^N\) in
a FOES model. For instance, this latter condition is fulfilled for all
linear exponential families from Section
\ref{discrete-exponential-family-models} (e.g.,
\eqref{eq:mod1}-\eqref{eq:mod2}) as well as all network models from Sections
\ref{restricted-boltzmann-machines}-\ref{deep-learning} (e.g.,
\eqref{eq:RBM1}-\eqref{eq:RBM2}). When parameters can be tuned in sign with
effects prescribed in the model condition PSR, unstable FOES models will
exhibit further probability sensitivities, as outlined in the following
extension of Theorem @ref(thm:degenFOES\}).

\BeginKnitrBlock{corollary}
\protect\hypertarget{cor:sign}{}{\label{cor:sign}}Let
\(P_{\boldsymbol \theta_N}\), with support \(\mathcal{X}^N\),
\(N\geq 1\), be a sequence of FOES models satisfying model condition
PSR. If the models \(P_{\boldsymbol \theta_N}\) are additionally
S-unstable, then
\begin{enumerate}[(i)]
\item the models  $P_{-\boldsymbol \theta_N}$  defined by $-\boldsymbol \theta_N$  are also S-unstable;
\item and for the complement $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c \equiv \mathcal{X}^N \setminus \mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ of any mode-set $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ under $\boldsymbol \theta_N$ from (\ref{eq:mode}), with $0<\epsilon<1$, it holds under $- \boldsymbol \theta_N$ that
    $$ 
    P_{-\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c) \rightarrow 1\qquad \mbox{as $N\to \infty$},
    $$
    while, by Theorem \ref{thm:degenFOES}, $P_{\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N} ) \rightarrow 1$ holds for $\boldsymbol X_N = (X_1,\ldots,X_N)$ under $\boldsymbol \theta_N$.
\end{enumerate}
\EndKnitrBlock{corollary}

@@

\BeginKnitrBlock{remark}
\iffalse <span class="remark"><em>Remark. \fi{} Model condition PSR and
Corollary \ref{cor:sign} can be extended to cases where parameter
components
\(\boldsymbol \theta_N =(\boldsymbol \theta_{1,N}, \boldsymbol \theta_{N,2})\)
are not all changed in sign (\(-\boldsymbol \theta_N\)) but, more
generally, are instead altered to another parameter configuration
\(\boldsymbol \theta_{N}^A = (\boldsymbol \theta_{N,1}^A, \boldsymbol \theta_{N,2}^A)\)
involving a switch in sign only among some model parameters
\(\boldsymbol \theta_{N,2}^A = -\boldsymbol \theta_{N,2}\) with
remaining parameters \(\boldsymbol \theta_{1,N}^A\) being arbitrarily
chosen. If a sign change occurs among parameters
(\(\pm \boldsymbol \theta_{N,2}\)) which dominate the probability
structure of the model, then an unstable model under
\(\boldsymbol \theta_N\) will then imply that many more unstable models
exist over a broad spectrum of possibilities for variations
\(\boldsymbol \theta_{N}^A\) of \(\boldsymbol \theta_N\).
\EndKnitrBlock{remark}

@@

As a baseline for comparisons, consider a simplistic model for
\(\boldsymbol X_N=(X_1,\ldots,X_N)\) with uniform probabilities over the
sample space, say
\(P_{\boldsymbol \theta_N}(\boldsymbol x)= |\mathcal{X}|^{-N}\),
\(\boldsymbol x \in \mathcal{X}^N\), where each random variable has
\(|\mathcal{X}| \geq 1\) outcomes. In contrast to instability, model
probabilities here are completely insensitive to changes in data
outcomes across the sample space, and the associated log-ratio of
extreme probabilities \eqref{eq:elpr} is \[
\frac{1}{N} \REP(\boldsymbol \theta_N)=0\quad \;(\mbox{uniform probability model}),
\] which is as small as possible. In fact, a LREP value of zero can only
occur for a FOES model having uniform probabilities, and such
equi-probability models are always S-stable.

@@

Due to their extreme sensitivity in probability structure, S-instability
in FOES models often translates to numerical complications, and in fact
obstructions, in both simulation and statistical inference based on
likelihoods.

@@

The rugged or volatile probability structure in unstable models can also
hamper efforts to maximize likelihood functions for statistical
inference. Essentially, when a FOES model is unstable along a parameter
sequence \(\boldsymbol \theta_N\), the same model can consequently
unstable along parameters \(-\boldsymbol \theta_N\) opposite from the
origin (model condition PSR and Corollary \ref{cor:sign}). This leads to
potential hyper-sensitivity in the likelihood function, which can be
pulled in extreme and opposite directions depending on how a parameter
lies with respect to zero and translates into numerical complications in
maximizing the objective function.

>>>>

\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{longtable,booktabs}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={A note on the instability and degeneracy of deep learning models},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{5}

@@

\bigskip
\bigskip
\bigskip
%\begin{center}
  %{\LARGE\bf Title}
  \title{\bf A note on the instability and degeneracy of deep learning models}
%\end{center}
\medskip

@@

We consider the behavior, and the potential impropriety, of sequences of
discrete probability models built to incorporate observations of
increasing sample size \(N\). Interest is in identifying instability in
such models, which is roughly characterized by probabilities with
extreme sensitivity to small changes in data configuration. The concept
of instability was introduced in the field of statistical physics (i.e.,
point processes) by Ruelle
(\protect\hyperlink{ref-ruelle1999statistical}{1999}) and then further
extended by Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) for a
family of exponential models. At issue, models exhibiting instability
are typically undesirable as these tend to provide poor representations
of data or data-generation. As an example, such models can include
near-degenerate distributions that assign all essential probability mass
to only a subset of an overall sample space. The latter issue in
connection to degeneracy has been recognized as a concern in that
dominant model outcomes may not resemble observed data (cf. Handcock
\protect\hyperlink{ref-handcock2003assessing}{2003}). As a compounding
issue, model instability often has direct negative impacts for
statistical inference and computations based on likelihood functions.
Namely, volatilities in probability structure can potentially hamper the
numerical evaluations required for maximum likelihood estimation as well
as other model-based simulations via Markov Chain Monte Carlo (MCMC).
These reasons motivate our general study of instability for a broad
class of probability models, described next.

@@

For random variables
\(\boldsymbol X \equiv\boldsymbol X_N= (X_1, \dots, X_N)\) with sample
space \(\mathcal{X}^N\), \(|\mathcal{X}| < \infty\), consider an
exponential family model for \(\boldsymbol X\) with probability mass
function given by
\begin{equation}
\label{eq:expo}
p_{N, \boldsymbol \theta}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) - \psi(\boldsymbol \theta)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
\end{equation}
depending on parameter vector
\(\boldsymbol \theta \in \Theta_N \subset \mathbb{R}^{k}\) and natural
parameter function
\(\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L\) with fixed
positive integers \(k\) and \(L\) denoting their dimensions. Above,
\(\boldsymbol g_N : \mathcal{X}^N \mapsto \mathbb{R}^L\) is a vector of
sufficient statistics, while \[
\psi(\boldsymbol \theta) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) \right], \qquad \boldsymbol \theta \in \Theta_N\equiv \{\boldsymbol \theta \in \mathbb{R}^k : \psi(\boldsymbol \theta) < \infty \},
\] denotes the normalizing function with parameter space \(\Theta_N\).
The natural parameter function \(\eta (\boldsymbol \theta)\) has a
linear form (i.e.,
\(\eta (\boldsymbol \theta)= \bm{A} \boldsymbol \theta\) for a given
\(L \times k\) matrix \(\bm{A}\)) in many common model formulations,
though may also be nonlinear (e.g., curved exponential families). In the
linear case, \(\eta (\boldsymbol \theta) = \boldsymbol \theta\) may be
generally assumed in the exponential parameterization with a minor
modification to the definition of sufficient statistics
\(\boldsymbol g_N(\boldsymbol x)\).

@@

For model sequences
\(P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \theta_N}(\boldsymbol x)\)
of the exponential type \eqref{eq:expo}, such as those in
\eqref{eq:mod1}-\eqref{eq:mod2}, note that the dimension \(k\) of the
parameter \(\boldsymbol \theta_N\in\Theta \subset \mathbb{R}^k\)
necessarily remains the same for all sample sizes \(N \geq 1\) as the
form of the natural parameter function \(\eta(\cdot)\) in \eqref{eq:expo}
and the number of sufficient statistics
\(\boldsymbol g_{N}(\boldsymbol x)\) do not depend on \(N\).
Consequently, \(\boldsymbol \theta_N\) lies in a parameter space of
fixed Euclidean dimension \(k\). However, this aspect need not be true
for other types of FOES models considered in Sections \ref{rbm} -
\ref{deep-learning}, where instead the numbers of parameters and
sufficient statistics commonly increase with the sample size \(N\).

\subsection{Restricted Boltzmann Machines}\label{rbm}

@@

This definition of S-unstable is a generalization or reinterpretation of
``unstable'' used in Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) by allowing
possibly non-exponential family models (e.g., RBM and DBM models in
Sections \ref{rbm}-\ref{deep-learning} as well as a potentially
increasing number \(q(N)\) of parameters through the parameter sequence
\(\boldsymbol \theta_N\in \mathbb{R}^{q(N)}\). While this definition
differs in form and scope from the original, it does match that in
Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) for the
special case of exponential models (cf.~Section
\ref{discrete-exponential-family-models} considered there. Section
\ref{illustrations} provides several examples of unstable models as well
as causes for model instability, where the latter may often be traced to
issues in model form (i.e., data functions) and/or parameterization. We
next describe several potentially undesirable features associated with
S-unstable FOES models.

\BeginKnitrBlock{remark}
\iffalse{} {Remark. } \fi{}In the definition \eqref{eq:Sun} of
S-instability, we note that the numerical measure
\(\REP(\boldsymbol \theta_N)/N\) of model instability is invariant to
\emph{independent replications} of data. That is, let \(M \geq 1\)
denote a possible number of replications and consider data
\(\boldsymbol Y_{N,M} \equiv (\boldsymbol X^{(1)}_N, \dots, \boldsymbol X^{(M)}_N)\)
formed by \(\{ \boldsymbol X^{(j)}_N\}_{j=1}^M\) as \(M\) iid
replications of a random vector \(\boldsymbol X_N=(X_1,\ldots,X_N)\),
where the latter follows a FOES model with probabilities
\(P_{ \boldsymbol \theta_N}(\boldsymbol x)>0\),
\(\boldsymbol x\in\mathcal{X}^N\). This leads to a joint model, say
\(P_{ \boldsymbol \theta_N}(\boldsymbol y)\),
\(\boldsymbol y\in\mathcal{X}^{NM}\), for \(\boldsymbol Y_{N,M}\)
consisting of \(N*M\) random variables in total. Then, the LREP for
\(\boldsymbol Y_{N,M}\), scaled by associated size, is given by
\begin{align*}
\frac{1}{NM}\REP_{\boldsymbol Y_{N,M}}( \boldsymbol\theta_N ) &\equiv \frac{1}{NM}\log\left[\frac{\max_{\boldsymbol y \in \mathcal{X}^{NM}}P_{ \boldsymbol \theta_N}(\boldsymbol y)  }{\min_{\boldsymbol y \in \mathcal{X}^{NM}}P_{ \boldsymbol \theta_N}(\boldsymbol y)} \right] \\
&= \frac{1}{NM}\log\left[\frac{\max_{\boldsymbol x \in \mathcal{X}^{N}}P_{ \boldsymbol \theta_N}(\boldsymbol x)  }{\min_{\boldsymbol x \in \mathcal{X}^{N}}P_{ \boldsymbol \theta_N}(\boldsymbol x)} \right]^M \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ),
\end{align*}
where
\(\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ) \equiv \REP( \boldsymbol\theta_N )\)
denotes the log-ratio of extremal probabilities for \(\boldsymbol X_N\)
defined from \eqref{eq:elpr}. That is, due to iid properties, the
sample-size corrected LREP for \(\boldsymbol Y_{N,M}\) equals the
analog, \(\REP(\boldsymbol\theta_N )/N\), from the underlying common
data model for \(\boldsymbol X_N\) alone, regardless of the level
\(M \geq 1\) of independent replication. Consequently, the definition of
an S-unstable model is unaffected by independent replication and all
instability properties may be characterized by those of one observation
from the common FOES model. For computational purposes, this aspect also
implies that if the original data \(\boldsymbol X_N=(X_1,\ldots,X_N)\)
in a FOES model consist of \(N\) iid random variables, then the
size-scaled log-ratio \eqref{eq:elpr} may be calculated as \[
\frac{1}{N}\REP( \boldsymbol\theta_N ) \equiv \frac{1}{N}\REP_{\boldsymbol X_{N}}( \boldsymbol\theta_N ) = \log\left[ \frac{\max_{ x \in \mathcal{X}}P_{\boldsymbol\theta_N}(X_1=x)} {\min_{ x \in \mathcal{X}}P_{\boldsymbol\theta_N}(X_1=x)} \right]
\] based on the extremal probabilities of just one random variable
\(X_1\).
\EndKnitrBlock{remark}

@@

The above model condition incorporates many standard parameterizations
and follows, for instance, whenever
\(P_{\boldsymbol \theta_N}(\boldsymbol x)/P_{\boldsymbol \theta_N}(\boldsymbol y) = [P_{-\boldsymbol \theta_N}(\boldsymbol x)/P_{-\boldsymbol \theta_N}(\boldsymbol y)]^{-1}\)
holds for outcomes \(\boldsymbol x, \boldsymbol y \in\mathcal{X}^N\) in
a FOES model. For instance, this latter condition is fulfilled for all
linear exponential families from Section
\ref{discrete-exponential-family-models} (e.g.,
\eqref{eq:mod1}-\eqref{eq:mod2}) as well as all network models from Sections
\ref{rbm}-\ref{deep-learning} (e.g., \eqref{eq:RBM1}-\eqref{eq:RBM2}). When
parameters can be tuned in sign with effects prescribed in the model
condition PSR, unstable FOES models will exhibit further probability
sensitivities, as outlined in the following extension of Theorem
@ref(thm:degenFOES\}).

\BeginKnitrBlock{corollary}
\protect\hypertarget{cor:sign}{}{\label{cor:sign} }Let
\(P_{\boldsymbol \theta_N}\), with support \(\mathcal{X}^N\),
\(N\geq 1\), be a sequence of FOES models satisfying model condition
PSR. If the models \(P_{\boldsymbol \theta_N}\) are additionally
S-unstable, then
\begin{enumerate}[(i)]
\item the models  $P_{-\boldsymbol \theta_N}$  defined by $-\boldsymbol \theta_N$  are also S-unstable;
\item and for the complement $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c \equiv \mathcal{X}^N \setminus \mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ of any mode-set $\mathcal{M}_{\epsilon, \boldsymbol \theta_N}$ under $\boldsymbol \theta_N$ from (\ref{eq:mode}), with $0<\epsilon<1$, it holds under $-\boldsymbol \theta_N$ that
    $$ 
    P_{-\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N}^c) \rightarrow 1\qquad \text{as } N\to \infty,
    $$
    while, by Theorem \ref{thm:degenFOES}, $P_{\boldsymbol \theta_N}( \boldsymbol X_N \in \mathcal{M}_{\epsilon, \boldsymbol \theta_N} ) \rightarrow 1$ holds for $\boldsymbol X_N = (X_1,\ldots,X_N)$ under $\boldsymbol \theta_N$.
\end{enumerate}
\EndKnitrBlock{corollary}

@@

\BeginKnitrBlock{remark}
\iffalse{} {Remark. } \fi{}Under the model condition PSR, Corollary
\ref{cor:sign} can also be extended to cases where parameter components
\(\boldsymbol \theta_N =(\boldsymbol \theta_{1,N}, \boldsymbol \theta_{N,2})\)
(say) are not all changed in sign (e.g., \(-\boldsymbol \theta_N\)) but,
more generally, are instead altered to another parameter configuration
\(\boldsymbol \theta_{N}^A = ( \boldsymbol \theta_{1,N}^A, \boldsymbol \theta_{2,N}^A )\)
involving a switch in sign only among some dominating model parameters
\(\boldsymbol \theta_{N,2}^A=- \boldsymbol \theta_{N,2}\) with remaining
parameters \(\boldsymbol \theta_{1,N}^A\) being arbitrarily chosen. If a
change sign occurs among parameters (\(\pm \boldsymbol \theta_{N,2}\))
which dominate the probability structure of the model, then the results
of Corollary \ref{cor:sign} can still hold with
\(\boldsymbol \theta_{N}^A\) replacing \(-\boldsymbol \theta_N\); as an
example of one sufficient condition, if
\(\lim_{N\to \infty} \max_{\boldsymbol x \in \mathcal{X}^N} |G_N(\boldsymbol x, -\boldsymbol \theta_{N}) - G_N(\boldsymbol x, \boldsymbol \theta_{N}^A)|=0\)
holds in addition to Corollary\textasciitilde{}1 assumptions, where \[
G_N(\boldsymbol x, \boldsymbol \theta) =   \frac{\log P_{\boldsymbol \theta}(\boldsymbol x) - \min_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y) }{\max_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y)  - \min_{\boldsymbol y \in\mathcal{X}^N}\log P_{\boldsymbol \theta}(\boldsymbol y) }, \quad \boldsymbol x \in\mathcal{X}^N,
\] represents a standardized form of \(\boldsymbol \theta\)-model
probabilities, then the results of Corollary \ref{cor:sign} apply to
\(\boldsymbol \theta_{N}^A\) in addition to \(-\boldsymbol \theta_N\).
As a consequence, an unstable model under \(\boldsymbol \theta_N\) can
then imply that many more unstable models exist over a broader spectrum
of possibilities for variations \(\boldsymbol \theta_{N}^A\) of
\(\boldsymbol \theta_N\), which involves some amount of sign change
among components of \(\boldsymbol \theta_N\).
\EndKnitrBlock{remark}

@@

As a baseline for comparisons, consider a simplistic model for
\(\boldsymbol X_N=(X_1,\ldots,X_N)\) with uniform probabilities over the
sample space, say
\(P_{\boldsymbol \theta_N}(\boldsymbol x)= |\mathcal{X}|^{-N}\),
\(\boldsymbol x \in \mathcal{X}^N\), where each random variable has
\(|\mathcal{X}| \geq 1\) outcomes. In contrast to instability, model
probabilities here are completely insensitive to changes in data
outcomes across the sample space, and the associated log-ratio of
extreme probabilities \eqref{eq:elpr} is \[
\frac{1}{N} \REP(\boldsymbol \theta_N)=0\quad \;(\text{uniform probability model}),
\] which is as small as possible. In fact, a LREP value of zero can only
occur for a FOES model having uniform probabilities, and such
equi-probability models are always S-stable.

@@

Due to their extreme sensitivity in probability structure, S-instability
in FOES models may often translate to numerical complications, and in
fact obstructions, in both simulation and statistical inference based on
likelihoods. We describe these aspects in Sections
\ref{mcmc}-\ref{bayes} with regard to data simulation, maximum
likelihood estimation and Bayes inference, respectively.

@@

Volatility in the probability structure of an unstable model can also
hamper efforts to maximize likelihood functions in statistical
inference. When a FOES model is unstable along a parameter sequence
\(\boldsymbol \theta_N\), the same model can further be unstable along
parameters \(-\boldsymbol \theta_N\) in an opposite direction from the
origin (model condition PSR and Corollary \ref{cor:sign}). This can
translate into potential sensitivity of the likelihood function around
zero, and lead to numerical complications in maximizing the objective
function. We next provide a discussion of this issue in a way that
builds upon and extends related findings by @Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}), who
largely focused on the case of one-parameter exponential models.
